\batcave: Calling somatic mutations with a tumor- and site-specific prior

### 2019-12-16 ResponseToReviews.tex

\section*{Reviewer 2}

Mannakee et al have developed a tool called BATCAVE for calling mutations from sequencing data. They claim to have used a novel approach of using tumor and site-specific priors for calling mutations. In my opinion authors have done a good job in terms of using a novel approach which harnesses patient/tumor specific information to call mutations. I agree with the authors that this approach was missing in other tools. I accept this study to be published given the authors address few points mentioned below:

1. Does BATCAVE accounts for tumor purity estimates while calling mutations?
\begin{response}
\RNG{Brian: How does MuTect handle this?}
\BKM{We do not. MuTect has a parameter for this, I will check exactly what it does.}
\end{response}

2. In case of multi-region sequencing data, used for studying spatial intra-tumor heterogeneity, usually the existing tools estimate very high levels of heterogeneity. I suppose this is because they are being very stringent in terms of calling mutations especially the ones with low vafs. In that case the people end up using other tools like bamreadcount etc. to check status of these mutations in the bam files and re-estimate the extent of heterogeneity. It will be important to see how BATCAVE performs in such datasets.
\begin{response}
We agree that multi-region data are increasingly important.
In fact, the breast cancer data we test on come from a multi-region study. 
Because we built BATCAVE on top of the single-sample caller MuTect, we called variants independently across the samples in that study, and BATCAVE performed well.
Unfortunately, the sequencing in that study is not deep enough to rigorously evaluate how estimates of heterogeneity would change with and without BATCAVE.

More broadly, the BATCAVE algorithm could be implemented in a multi-regional caller that takes explicit advantage of the relationship between samples.
But such an implementation is beyond the scope of the current manuscript.

We have added a new paragraph in Discussion regarding multi-region samples, to address these issues.
\end{response}

### 2019-12-16 NAR_DRAFT.tex
Address multi-region comments

\rngcomment{
Following MuTect \cite{Cibulskis2013}, \batcave assumes a uniform prior on the variant allele frequency.
In general, low-frequency variants are expected to be more common than high-frequency variants.
Nevertheless, the uniform frequency prior is appropriate when the goal is to make the best call possible at each potential variant site.
By favoring low-frequency and thus typically low-evidence variants, a nonuniform prior would increase the false positive rate.
Note however, that if the goal were not to best identify individual variants, but rather to best estimate the distribution of variant allele frequencies within the tumor, then a nonuniform frequency prior would be appropriate \cite{Nielsen2012}.
}

@@

@@

\batcave divides the data into two classes: high- and low-confidence variants.
The high-confidence variants are used to estimate the mutational prior and mutation rate, which are then used to improve the calling of low-confidence variants.
Statistically, this is an empirical Bayesian approach~\cite{Robbins1954}, in which the high and low-confidence variants are treated as parallel experiments \cite{Morris1983,Efron2014}. 
In general, high-confidence variants tend to have relatively high allelic frequencies, and consequently tend to have arisen early in tumor development.
An  implicit assumption of our approach is that the mutational process does not change between high- and low-confidence variants, implying that the mutational profile of the tumor is temporally constant.
Recent studies have found differences in mutational profiles among variants of different allelic frequencies~\cite{Rubanova2018a}, and although in general those differences are relatively small, they can be significant in some cases.
\bkmcomment{We provide the user with the ability to output a file with the contents of the estimated mutation profile as each high-confidence variant is evaluated.
In cases where the user suspects a shift in mutational process, we encourage using this file to verify that the final mutation profile is representative of the processes active at low frequency.}
A potential extension of the \batcave algorithm is to process potential variants in order of descending allelic frequency and to update the estimated mutational prior as the algorithm proceeds.
This approach might increase sensitivity to low-frequency variants generated by recently-arisen mutational processes, at the cost of potentially increasing sensitivity to patterns of sequencing error.
\bkmcomment{We use properties of the allele frequency distribution generated by mutation acquisition during exponential growth.
Copy number alterations can distort the allele frequency distribution.
While copy number alterations in a large number of genomic regions would be required to significantly affect the estimate of mutation rate, an additional extension of the \batcave algorithm could be to incorporate external copy number calls in order to adjust allele frequencies.}
Our results show that adding a mutational prior substantially improves probabilistic variant calling, particularly for tumors with concentrated profiles.
Improved variant calling increases the benefit-to-cost ratio of deep sequencing in both research and clinical applications.
Moreover, \batcave proves to be a better calibrated caller than vanilla MuTect (Fig.~\ref{NAR-ppv_fig}).
Different users will prefer different tradeoffs in terms of precision and recall, which can be more accurately made with \batcave.
Our R implementation, \texttt{batcaver}, can be easily incorporated into any MuTect-based pipeline, and the mutational profile algorithm can be incorporated into many other callers.

>>>>

\rngcomment{
Following MuTect \cite{Cibulskis2013}, \batcave assumes a uniform prior on the variant allele frequency.
In general, low-frequency variants are expected to be more common than high-frequency variants.
Nevertheless, the uniform frequency prior is appropriate when the goal is to make the best call possible at each potential variant site.
By favoring low-frequency and thus typically low-evidence variants, a nonuniform prior would increase the false positive rate.
Note however, that if the goal were not to best identify individual variants, but rather to best estimate the distribution of variant allele frequencies within the tumor, then a nonuniform frequency prior would be appropriate~\cite{Nielsen2012}.
}

@@

\rngcomment{
Sequencing multiple regions of the same tumor is of growing interest for understanding tumor heterogeneity and evolution \cite{McGranahan2017}.
In fact, the breast cancer data we analyze came from a multi-region study~\cite{Shi2018}.
We called variants independently across samples, because MuTect only considers a single sample at a time.
But calling variants jointly across samples could substantially increase sensitivity.
In the future, the \batcave algorithm could be integrated into a multi-sample caller such as MultiSNV~\cite{Josephidou2015}.
In that case, an optimal implementation would likely exploit the relationship between mutational priors in related samples.
}

@@

\batcave divides the data into two classes: high- and low-confidence variants.
The high-confidence variants are used to estimate the mutational prior and mutation rate, which are then used to improve the calling of low-confidence variants.
Statistically, this is an empirical Bayesian approach~\cite{Robbins1954}, in which the high and low-confidence variants are treated as parallel experiments \cite{Morris1983,Efron2014}. 
In general, high-confidence variants tend to have relatively high allelic frequencies, and consequently tend to have arisen early in tumor development.
An  implicit assumption of our approach is that the mutational process does not change between high- and low-confidence variants, implying that the mutational profile of the tumor is temporally constant.
Recent studies have found differences in mutational profiles among variants of different allelic frequencies~\cite{Rubanova2018a}, and although in general those differences are relatively small, they can be significant in some cases.
\bkmcomment{We provide the user with the ability to output a file with the contents of the estimated mutation profile as each high-confidence variant is evaluated.
In cases where the user suspects a shift in mutational process, we encourage using this file to verify that the final mutation profile is representative of the processes active at low frequency.}
A potential extension of the \batcave algorithm is to process potential variants in order of descending allelic frequency and to update the estimated mutational prior as the algorithm proceeds.
This approach might increase sensitivity to low-frequency variants generated by recently-arisen mutational processes, at the cost of potentially increasing sensitivity to patterns of sequencing error.

Our results show that adding a mutational prior substantially improves probabilistic variant calling, particularly for tumors with concentrated profiles.
Improved variant calling increases the benefit-to-cost ratio of deep sequencing in both research and clinical applications.
Moreover, \batcave proves to be a better calibrated caller than vanilla MuTect (Fig.~\ref{NAR-ppv_fig}).
Different users will prefer different tradeoffs in terms of precision and recall, which can be more accurately made with \batcave.
Our R implementation, \texttt{batcaver}, can be easily incorporated into any MuTect-based pipeline, and the mutational profile algorithm can be incorporated into many other callers.
